Extended early adulthoods

### 2023-02-21 draft.Rmd
Changes some wording in the main draft. Plausible structure for the extended abstract

---
title: "Extended early adulthoods"
author: "Kevin Kiley & Nicolas Restrepo"
fontsize: 12pt
output:
  pdf_document: default
  html_document: default
header-includes:
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage{lineno}
  - \linenumbers
  - \def\linenumberfont{\normalfont\tiny\sffamily}
bibliography: efw_bib.bib
abstract: "Young adulthood is consistently identified as a “formative period” in the life course when people have heightened probability of attitude change. This probability has often been attributed to cognitive factors like brain plasticity as well as the social conditions of adolescence. Over the past 70 years, changes in the timing of major life course events such as marriage have transformed and extended this life stage. We leverage these changes to explore whether these life course changes, as well as broader social changes, have affected how likely people are to make durable changes of attitudes. We find strikingly little difference in the opinion behaviors among young people in different eras. People under 30 are similarly different from people over 30 in all eras, with slightly less stable attitudes. The overall pattern suggests either that social conditions are not strongly related to attitude stability or that different social shifts have had counter-vailing effects that produce aggregate stability in rates." 
editor_options: 
  markdown: 
    wrap: sentence
---

@@

Adolescence and young adulthood are typically identified in social science literature as "impressionable years" -- a stage of the life course in which people are more susceptible to broad social currents, when their opinions are more malleable, and when peoples' core ideas and attitudes are formed [].
Research regularly finds that people are more likely to change their attitudes and opinions in these life stages than other life stages, with opinions becoming more stable once these life stages have passed [].
This pattern, in which young people are open to attitude change and become resistant over time, underpins dynamics around cohortization that are frequently invoked in larger models of social change [@ryder1965; @mannheim1952; @vaisey2016].

@@

Clarifying the mechanism that drives the frequency of opinion change during this formative period is important for understanding the broader process of social change and cultural evolution.
If openness to change in early adulthood is principally a function of the social structure of life in that time period, changes in the timing and ubiquity of life course events like marriage and childbearing can have significant impacts on the broader process of social and cultural change [].
Similarly, social changes such as the deinstitutionalization of marriage [@cherlin2004; @cherlin2020; @robbins2022] that decrease distinctions between life course stages might make older people more open to change in more recent eras than in previous eras, diminishing the importance of "cohorts" for social change.
If the mechanisms are principally cognitive and biological, rather than social, then these changes are unlikely to affect rates of social and cultural change in the same way.

Over the course of the 20th century, the timing of life-course transitions such as completing schooling, entering the workforce, getting married, leaving one's parents' home, and having children, have changed dramatically, with most of these event occurring later in the life course in more recent cohorts [@buchmann1989; @bruckner2005].
Similarly, this stage of life has become more detached from family and other role obligations, giving people more freedom to explore new ideas and lifestyle choices [@rosenfeld2007].
This suggests both a lengthening and a intensification of this stage of the life course in ways that should increase the rates at which people make changes of opinion.
In this paper, we attempt to clarify the relative contributions of these different mechanisms by exploring how these changes have affected rates of change in political attitudes over time, especially differences in rates of opinion change between young adults and the rest of the population.
We compare rates of opinion change in comparable general political attitudes across four eras -- 1956-1960; 1972-1976; 1992-1996; and 2006-2010 -- and between young people and old people in these eras to evaluate a range of competing hypotheses about what accounts for the divergence in attitude behaviors over the life course.

@@

The idea that the early stages of the life course are especially influential in the process of attitudinal formation is dominant across contemporary accounts of socialization [@dinas2014; @guhin2021].
The argument -- at times implicit -- is that the circumstances in which people are brought up leave an enduring and traceable mark on their worldviews (Gerber and Green 1998).
In the study of political socialization, this perspective enjoys a longstanding history.
"Generational effects" are used to explain the development of political preferences, longitudinal shifts in attitudes, and differences across age-groups [@mannheim1952; @bartels2014].
However, the notion of "impressionable years" is more general.
The literature on cohort effects rests on the assumption that our early years play key role in shaping not just political opinions, but peoples' cognition on many dimensions (Elder 2018).
Peoples' early years will leave a mark that will be predictive of their life outcomes, well beyond when those impressionable years have passed.
Thus, the notion of a formative period -- at the early stages of the life-course -- underpins diverse and longstanding lines of research.

The "impressionable years" hypothesis -- that people demonstrate more opinion change early in life followed by a period of relative stability, especially in middle age -- has often been tested against other theories of opinion change over the life course.
The most common are an aging stability or increasing persistence hypothesis that argues that openness to change gradually declines over time and a life stages hypothesis, in which people are open to attitude change both early in life and later in life, with stability principally happening in middle ages [@visser2004a].
While different attitudes occasionally follow slightly different patterns, the impressionable years hypothesis or aging stability hypothesis has generally been shown to be true for broad political ideology, partisan identification, and specific policy preferences [@krosnick1989; @alwin1991; @sears1999]; social attitudes [@danigelis2007]; and interest in politics [@prior2010]; as well as things like personality [@wagner2019].
More generally, recent research has found that, with a couple notable exceptions of high-salience events such as national shifts around gay rights, durable change of social opinion is quite rare among U.S. adults [@kiley2020], and Vaisey and Lizardo [-@vaisey2016] find that cohort explains much more variance than period for most items in the General Social Survey, suggesting a strong impressionable years pattern.

@@

Second, people in the "impressionable years" window spend their time in different social contexts than people in middle age.
Data from the American Time Use Study shows that people under age 25 spend a larger portion of the day on leisure time and educational activities and a smaller part of the day doing work and work-related activities than people between the ages of 25 and 55.
These contexts, especially education, are explicity designed to expose people to new information in ways that can lead to updated opinions.
For example, there is some evidence that college attendance leads to changes in opinions about civil liberties, gender egalitarianism, and racial inequality [@campbell2016; @wodtke2018].
Social networks also differ in this time period.
Previous work has found that young peoples' networks tend to be larger and include more friends than kin, for example [@marsden1987].
It could be the case that larger "weak tie" networks expose younger people to more new information that drives opinion change [@granovetter1975; @burt2005].

@@

Similarly, mid-life is associated with more homogeneous networks [@marsden1985; @diprete2011], which experimental research suggests reduces the propensity to adopt new beliefs [@vissermirible].
If young people exist in less dense, more heterogeneous social networks, they might it easier to update their networks to reflect new beliefs rather than constrain their beliefs to their social networks [], or close off parts of their lives to people who might not be supportive of new attitudes.

@@

Third, we can distinguish between the "elongation of young adulthood" and the "intensification of young adulthood" arguments by exploring the associaiton of life-course transitions and opinion change across cohorts.
The elongation of young adulthood argument principally suggests that delaying the life course transitions is what would explain heightened rates of attitudinal updating in young people in recent cohorts.
In contrast, the "intensification of young adulthood" argument suggests that attitudinal updating may be occurring at heightened rates in recent cohorts net of these life course transitions.

@@

The four panels were highly inconsistent in how they measured whether people had children.
The 1956-60 panel of the ANES asked respondents how many children under 18 years old they had in the household.
We coded individuals as having children in the first wave of the panel if they answered at least one to this question.
The 1970s panel of the ANES asked respondents in the first wave if they had any school-aged children.
In the second wave, they asked respondents if they ever had children, how many, and the ages of all children in the household.
We code respondents as "having children in the first wave of the panel if they reported having at least one child older than one in the second wave of the panel. Because some of these questions focus on ever having children while others focus on children in the household, we caution readers from making any significant judgments about these coefficients.

@@

The broadest challenge for our analysis is that it is not obvious what it means for a person to demonstrate opinion change in a survey.
Debates about how much "change" we see in opinions often focus on differences in the approach to measurement [@converse1965; @achen1975; @ansolabehere2008].
The kinds of questions we seek to examine, general political dispositions, tend to be asked on five- or seven-point ordinal scales, either from strongly agree to strongly disagree, or from one policy position ("government should help minority groups") to another ("minority groups should help themselves").
These kinds of scales support a wide range of different measures of opinion change and stability.
Rather than adopt a single approach to quantify the rates of change across questions, we use a variety of approaches with different assumptions about what "change" entails and attempt to identify patterns that are robust across a range of assumptions.
When these approaches disagree, having multiple measures allows us to more clearly evaluate what forms of change might be driving differences between groups.

@@

>>>>

---
title: "Extended early adulthoods"
author: "Kevin Kiley & Nicolas Restrepo"
fontsize: 12pt
output:
  pdf_document: default
  html_document: default
header-includes:
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage{lineno}
  - \linenumbers
  - \def\linenumberfont{\normalfont\tiny\sffamily}
bibliography: efw_bib.bib
abstract: "Young adulthood is consistently identified as a “formative period” in the life course when people have heightened probability of attitude change. This probability has often been attributed to cognitive factors like brain plasticity as well as the social conditions of adolescence. Over the past 70 years, changes in the timing of major life course events such as marriage have transformed and extended this life stage. We take advantage of these changes to explore whether they -- as well as broader social changes -- have affected how likely people are to make durable changes of attitudes. We find strikingly little difference in the opinion behaviors among young people in different eras. People under 30 are similarly different from people over 30 in all eras, with slightly less stable attitudes. The overall pattern suggests either that social conditions are not strongly related to attitude stability or that different social shifts have had counter-vailing effects that produce aggregate stability in rates." 
editor_options: 
  markdown: 
    wrap: sentence
---

@@

Adolescence and young adulthood are typically identified in the social science literature as "impressionable years" -- a stage of the life course in which people are more susceptible to broad social currents, when their opinions are more malleable, and when peoples' core ideas and attitudes are formed [].
Research regularly finds that people are more likely to change their attitudes and opinions in these life stages than other periods, with opinions becoming more stable once these stages have passed [].
This pattern, in which young people are open to attitude change and become resistant over time, underpins dynamics around cohortization that are frequently invoked in larger models of social change [@ryder1965; @mannheim1952; @vaisey2016].

@@

Clarifying the mechanism that drives the frequency of opinion change during this formative period is important for understanding the broader process of social change and cultural evolution.
If openness to change in early adulthood is principally a function of the social structure of life in that time period, changes in the timing and ubiquity of life course events like marriage and childbearing can have significant impacts on the broader process of social and cultural change. Similarly, social changes such as the deinstitutionalization of marriage [@cherlin2004; @cherlin2021; @brückner2005] that decrease distinctions between life course stages might make older people more open to change in more recent eras than in previous eras, diminishing the importance of "cohorts" for social change.
If the mechanisms are principally cognitive and biological, rather than social, then these changes are unlikely to affect rates of social and cultural change in the same way.

Over the course of the 20th century, the timing of life-course transitions such as completing schooling, entering the workforce, getting married, leaving one's parents' home, and having children, have changed dramatically, with most of these event occurring later in the life course in more recent cohorts [@buchmann1989; @brückner2005].
Similarly, this stage of life has become more detached from family and other role obligations, giving people more freedom to explore new ideas and lifestyle choices [@rosenfeld2005].
This suggests both a lengthening and a intensification of this stage of the life course in ways that should increase the rates at which people make changes of opinion.
In this paper, we attempt to clarify the relative contributions of these different mechanisms by exploring how these changes have affected rates of change in political attitudes over time, especially differences in rates of opinion change between young adults and the rest of the population.
We compare rates of opinion change in comparable general political attitudes across four eras -- 1956-1960; 1972-1976; 1992-1996; and 2006-2010 -- and between young people and old people in these eras to evaluate a range of competing hypotheses about what accounts for the divergence in attitude behaviors over the life course.

@@

The idea that the early stages of the life course are especially influential in the process of attitudinal formation is dominant across contemporary accounts of socialization [@dinas2014; @guhin2020].
The argument -- at times implicit -- is that the circumstances in which people are brought up leave an enduring and traceable mark on their worldviews (Gerber and Green 1998).
In the study of political socialization, this perspective enjoys a longstanding history.
"Generational effects" are used to explain the development of political preferences, longitudinal shifts in attitudes, and differences across age-groups [@mannheim1952; @bartels2014].
However, the notion of "impressionable years" is more general.
The literature on cohort effects rests on the assumption that our early years play key role in shaping not just political opinions, but peoples' cognition on many dimensions (Elder 2018).
Peoples' early years will leave a mark that will be predictive of their life outcomes, well beyond when those impressionable years have passed.
Thus, the notion of a formative period -- at the early stages of the life-course -- underpins diverse and longstanding lines of research.

The "impressionable years" hypothesis -- that people demonstrate more opinion change early in life followed by a period of relative stability, especially in middle age -- has often been tested against other theories of opinion change over the life course.
The most common are an aging stability or increasing persistence hypothesis that argues that openness to change gradually declines over time and a life stages hypothesis, in which people are open to attitude change both early in life and later in life, with stability principally happening in middle ages [@visser2004].
While different attitudes occasionally follow slightly different patterns, the impressionable years hypothesis or aging stability hypothesis has generally been shown to be true for broad political ideology, partisan identification, and specific policy preferences [@krosnick1989; @alwin1991; @sears1999]; social attitudes [@danigelis2007]; and interest in politics [@prior2010]; as well as things like personality [@wagner2019].
More generally, recent research has found that, with a couple notable exceptions of high-salience events such as national shifts around gay rights, durable change of social opinion is quite rare among U.S. adults. [@kiley2020], and Vaisey and Lizardo [-@vaisey2016] find that cohort explains much more variance than period for most items in the General Social Survey, suggesting a strong impressionable years pattern.

@@

Second, people in the "impressionable years" window spend their time in different social contexts than people in middle age.
Data from the American Time Use Study shows that people under age 25 spend a larger portion of the day on leisure time and educational activities and a smaller part of the day doing work and work-related activities than people between the ages of 25 and 55.
These contexts, especially education, are explicitly designed to expose people to new information in ways that can lead to updated opinions.
For example, there is some evidence that college attendance leads to changes in opinions about civil liberties, gender egalitarianism, and racial inequality [@campbell2016; @wodtke2018].
Social networks also differ in this time period.
Previous work has found that young peoples' networks tend to be larger and include more friends than kin, for example [@marsden1987].
It could be the case that larger "weak tie" networks expose younger people to more new information that drives opinion change [@granovetter1973; @burt2005].

@@

Similarly, mid-life is associated with more homogeneous networks [@marsden1987; @diprete2011], which experimental research suggests reduces the propensity to adopt new beliefs [@visser2004].
If young people exist in less dense, more heterogeneous social networks, they might it easier to update their networks to reflect new beliefs rather than constrain their beliefs to their social networks or close off parts of their lives to people who might not be supportive of new attitudes.

@@

Third, we can distinguish between the "elongation of young adulthood" and the "intensification of young adulthood" arguments by exploring the association of life-course transitions and opinion change across cohorts.
The elongation of young adulthood argument principally suggests that delaying the life course transitions is what would explain heightened rates of attitudinal updating in young people in recent cohorts.
In contrast, the "intensification of young adulthood" argument suggests that attitudinal updating may be occurring at heightened rates in recent cohorts net of these life course transitions.

@@

The four panels were highly inconsistent in how they measured whether people had children.
The 1956-60 panel of the ANES asked respondents how many children under 18 years old they had in the household.
We coded individuals as having children in the first wave of the panel if they answered at least one to this question.
The 1970s panel of the ANES asked respondents in the first wave if they had any school-aged children.
In the second wave, they asked respondents if they ever had children, how many, and the ages of all children in the household.
We code respondents as "having children" in the first wave of the panel if they reported having at least one child older than one in the second wave of the panel. Because some of these questions focus on ever having children while others focus on children in the household, we caution readers from making any significant judgments about these coefficients.

@@

The broadest challenge for our analysis is that it is not obvious what it means for a person to demonstrate opinion change in a survey.
Debates about how much "change" we see in opinions often focus on differences in the approach to measurement [@converse1964; @achen1975; @ansolabehere2008].
The kinds of questions we seek to examine, general political dispositions, tend to be asked on five- or seven-point ordinal scales, either from strongly agree to strongly disagree, or from one policy position ("government should help minority groups") to another ("minority groups should help themselves").
These kinds of scales support a wide range of different measures of opinion change and stability.
Rather than adopt a single approach to quantify the rates of change across questions, we use a variety of approaches with different assumptions about what "change" entails and attempt to identify patterns that are robust across a range of assumptions.
When these approaches disagree, having multiple measures allows us to more clearly evaluate what forms of change might be driving differences between groups.

@@

## References
